Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-09-23 analysis/paper/paper.qmd
fix everything from SP except table format (do manually), Supplementary Appendix (no need)

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Spheroids are widely distributed lithic artifacts, yet their production strategies and functions remain debated. Traditional morphology-based typologies obscure their technological features and variability. Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a method of quantitative morphological analysis, integrated with technological measures, to assess production strategies. Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates that spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy. Our results further demonstrate that spheroids can be clearly distinguished from polyhedrons and multifacial cores exhausted through reduction. Our methods for investigating spheroid production strategies reveal a higher level of technological complexity than previously assumed for early lithic industries in northern China. Our study also provides a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production but also more broadly lithic technological research for any period and region.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
#| label: fig-QSY-geography-chronology
#| fig-cap: "Geological Context and Chrono-Stratigraphy of the Qianshangying site complex)"

@@

In terms of dimensions, the average length, width, thickness, and mass of spheroids subspheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics. few spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. Polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files spheroids_metrics.xlsx). However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement. Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid. This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form. Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].

@@

This work was financially supported through the National Natural Foundation of China (Grant No. 42371165), Please list foundation from Ben, and China Scholarship Council Programme for international student (CSC No. 202404910542)

>>>>

title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
crossref:
  fig-title: "Fig."
  fig-prefix: "Fig."
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Spheroids are widely distributed lithic artifacts, yet their production strategies and functions remain debated. Traditional morphology-based typologies obscure their technological features and variability. Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a method of quantitative morphological analysis, integrated with technological measures, to assess production strategies. Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates that spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy. Our results further demonstrate that spheroids can be clearly distinguished from polyhedrons and multifacial cores exhausted through reduction. Our methods for investigating spheroid production strategies reveal a higher level of technological complexity than previously assumed for early lithic industries in northern China. Our study also provides a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production but also more broadly lithic technological research for any period and region.
editor_options: 
  chunk_output_type: console
---

@@

```{r, echo=FALSE}
#| label: fig-QSY-geography-chronology
#| fig-cap: "Geological Context and Chrono-Stratigraphy of the Qianshangying site complex): A-B) Geographic location of Nihewan basin. C) Google Earth image showing major sites in the Jijiazhuang Platform. D)  Landscape and different localities of Qianshangying site complex. E) Archaeostratigraphy sequence and ESR dating results of Qianshangyinig site."

@@

In terms of dimensions, the average length, width, thickness, and mass of spheroids subspheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics. Few spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. Polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files QSY_A.xlsx and QSY_B.xlsx). However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement. Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid. This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form. Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].

@@

This work was financially supported through the National Natural Foundation of China (Grant No. 42371165), and China Scholarship Council Programme for international student (CSC No. 202404910542)
